\documentclass[DM,authoryear,toc]{lsstdoc}
\input{meta}

% Package imports go here.

% Local commands go here.

%If you want glossaries
%\input{aglossary.tex}
%\makeglossaries

\title{Options for Alert Packets}

% Optional subtitle
% \setDocSubtitle{A subtitle}

\author{%
Melissa Graham and the Data Management System Science Team
}

\setDocRef{DMTN-248}
\setDocUpstreamLocation{\url{https://github.com/lsst-dm/dmtn-248}}

\date{\vcsDate}

% Optional: name of the document's curator
% \setDocCurator{The Curator of this Document}

\setDocAbstract{%
A review of the scientific impact of options for alert packets, with recommendations.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{1}{2023-01-23}{First version.}{Melissa Graham}
}


\begin{document}

% Create the title page.
\maketitle
% Frequently for a technote we do not want a title page  uncomment this to remove the title page and changelog.
% use \mkshorttitle to remove the extra pages


\section{Introduction}\label{sec:introduction}

Alerts will play a major role in the scientific impact of the LSST.
They are the only data product that is both world public (exempt from any proprietary period) and publicly accessible,
the latter meaning that they are distributed to community brokers and not solely available via the Rubin Science Platform (RSP).

The alert stream and its packets' contents are described in the Data Products Definitions Document (DPDD; \citeds{LSE-163}).
The motivation behind their design is to rapidly distribute all of the LSST data about a difference-image detection which a broker (or their users) would need to assess, classify, and prioritize the alert for follow-up observations within minutes.

This document considers options for the alert stream and alert packet contents to maximize science.
Most options look at either reducing the packet size, or improving the scientific utility of packet contents.

\textbf{Reduce packet size}:
Some options below aim to reduce the size of the alert packet, potentially enabling wider full-stream redistribution to downstream brokers or avoiding bottlenecks in alert distribution for crowded fields.
The size of an alert packet is estimated to be 82 KB, based on simulations described in Section 3.5 of the DPDD (see also the Alerts Key Numbers document; \citeds{DMTN-102}).
Reducing packet size was more of a concern in the earlier days of construction, before the decision to support seven full-stream brokers was made.

\textbf{Improve scientific utility}:
Other options below aim to improve the scientific utility of alert packet contents by providind additional or differently-formatted data.

This document also includes a brief discussion about the science impacts for the various ways of dealing with delayed alerts when, e.g., there are $>$40000 alerts per visit (Section \ref{sec:delayed}).


\section{Remove historical records}\label{sec:histories}

Alert packets will include ``\textit{any DiaSource and DiaForcedSource records that exist, and difference image noise
estimates where they do not, taken from the previous 12 months}" (Section 3.5.1 of the DPDD).
The history would account for about 27 KB, or about 33\% of the total alert packet size.

The motivation for including historical records in alert packets is to enable brokers to (re-)assess the \emph{full} time-domain event in order to make a robust classification (or prioritization) for follow-up.

The proposal is that instead of including the history in every packet, brokers could retrieve it only the fraction of the alerts for which it is needed.

Brokers (and their users) would still need the list of the unique identifers for the DiaSource and DiaForcedSource records associated with the same DiaOject as the alert-triggering DiaSource, even if their full records were not included in the alert packet.
Removing this list is \emph{not} being considered, for the following reasons:

\begin{enumerate}
\item The LSST Prompt Processing pipeline associates DiaSources into DiaObjects, and it would be challenging, time-consuming, and redundant to have brokers repeat this for all new alerts.
\item The association of DiaSources into DiaObjects is probabilistic, and in rare cases might change over time (e.g., strongly-lensed supernovae, or line-of-sight transient superpositions, where two distinct difference-image sources are blended/separated in poor/good seeing). In such cases the set of all DiaSource records which compose the history might change.
\end{enumerate}

Brokers that save all past alerts in their own archives would be able to retrieve the full records for associated DiaSources locally (but not DiaForcedSources, as they would never have been in an alert packet).
Brokers with authenticated RSP access would be able to query the PPDB to retrieve the full records for associated DiaSources and DiaForcedSources.
Brokers that do not save all past alerts and/or do not have authenticated RSP access to the PPDB would have no historical information aside from the number of associated DiaSources (i.e., the number of past detections).

At best, the lack of historical records in alerts would add a layer of complication and a potential delay to the brokers' functionailty, and inhibit the scientific assessment of time-domain events.
At worst, it would be a severe risk to brokers' ability to process alerts.

\textbf{Recommendation}: Do not remove historial records, it imposes a risk to time-domain science.


\section{Remove postage stamps}\label{sec:stamps}

The postage stamps are, at minimum, 6$\times$6 arcseconds (30$\times$30 pixels) and contain flux, variance, and mask extensions for both the template and difference image, plus a header of metadata (DPDD).
The stamps would accounts for about 18 KB, or about 20\% of the total alert packet size.

The motivation for including postage stamps in the alert packets is to enable brokers to use image-based machine learning tools (e.g., custom real/bogus scores; host+supernova classifiers), 2-dimenstional flux distribution (e.g., trailed sources; cometary outbursts), or environmental context (e.g., field crowdedness) to classify or prioritze alerts.

The proposal is to remove image stamps from the alert packets, in favor of one of the two options below, with which brokers could instead retrieve the postage stamps for only those alerts for which it is needed (e.g., unclassified time-domain events).

\begin{enumerate}
\item \textbf{Image cutout service}. Brokers with authenticated RSP access could use the image cutout service to create and retrieve stamps.
\item \textbf{URL to the postage stamp}. A URL to could be put into the alert instead of the postage stamp which points to an automated public cutout service or the pre-made postage stamps on a server.
\end{enumerate}

The 80 hour embargo on all new images and difference images renders the option to use an image cutout service non-viable (regardless of whether it is via the RSP or a public service).

The option for Rubin Observatory to create and maintain a public server of pre-made postage stamps has the advantage that brokers (or downstream brokers) do not need to save the stamps, and can retrieve them at any time.

However, this option also has two main drawbacks: it is not currently planned for development, and it could impose a delay on brokers' alert processing and analysis.
The latter is unlikely to cause a bottleneck whe multiple brokers attempt to retrieve large numbers of stamps simulataneously.
For example, if 5 brokers retrieve 500 stamp sets per visit, that's 2500 stamp sets every 35 seconds, and at 18 KB per stamp set the data rate would be 10.5 Mbps (5\% of one full alert stream).

\textbf{Recommendation}: Do not remove postage stamps, it imposes a risk to time-domain science.


\section{Provide large multi-resolution postage stamps}\label{sec:bigstamps}

The proposal is to provide larger multi-resolution postage stamps in the alert packets by binning the pixels with a bin size that increases towards the edges of the images, such that several arcminutes can be included.

The motivation for this proposal is to enable time-domain science that relies on the rapid association of low-redshift transients with their wide-area host galaxies, using algorithms such as ``DELIGHT: Deep Learning Identification of Galaxy Hosts of Transients using Multiresolution Images" (Förster et al. 2022).

The default 6$\times$6 arcsecond postage stamp size is large enough to contain the angular diameter (30 kpc) of a Milky Way-like galaxy at a redshift of 0.5, which will be a fairly typical supernova host galaxy in the LSST data set.
The fraction of all alerts that will be low-redshift transients with wide-area host galaxies will be $<$1\% \citedsp{DMTN-102}.

Host association will already be included in the alert packet as part of the DiaObject, as described in \citeds{DMTN-151} and in Table 3 of the DPDD, including potential nearby extended and low-redshift galaxies (see columns nearbyExtObj and nearbyLowzGal).

This type of host-galaxy identification could be done by accessing the template images (which are not subject to the 80 hour embargo) via the image cutout service.

\textbf{Recommendation}: At this time, do not increase the size of postage stamps.


\section{Compress alert packets}\label{sec:compression}

The application of gzip compression could further reduce the size of a full alert to 65 KB (80\%).
This might help to avoid alert distribution bottlenecks, lower the cost for brokers' storage needs, and potentially enable additional full streams in the future -- all of which could positively impact time-domain science with the LSST.

However, the time and computational resources required to compress the alert packets needs to be considered.
For example, gzip compression at 50 MB/s to compress 10000 alerts could take 10000 $\times$ 0.08 MB per alert, 50 MB/s, or 16 seconds, which is very significant in light of the 60 second alert distribution timescale. 

Furthermore, compressing the alert packets forces brokers to then decompress the alerts on arrival, which would incur further delay.
Science goals requiring very low-latency alerts distribution might be negatively impacted by compression.

Compression should only be consisdered if an algorithm could provide a compression rate of 10 MB/s, requiring only a few seconds to compress a visit's worth of alerts, or if the alert distribution latency requirement is relaxed from 1 minute.

\textbf{Recommendation}: Do not compress alert packets, it imposes a risk to rapid time-domain science.


\section{Enable multiple packet formats}\label{sec:multiformats}

The proposal is to allow brokers to specify whether their stream should have histories or postage stamps removed.

The main motivation here is not scientific, but to help reduce brokers' processing costs spent removing unneeded information from alert packets. 

There are a few drawbacks to this proposal, but they are mostly technical:

\begin{enumerate}
\item New scope, as the current plan is for Prompt Processing to create a uniform alert packet.
\item Streams of full-sized packets could unfairly suffer delays more often (e.g., in crowded fields).
\item Non-identical packets might cause bookkeeping issues for downstream brokers subscribing to multiple brokers.
\end{enumerate}

These drawbacks could impose a risk to the particular science goals of brokers using full-sized or non-identical alerts.

\textbf{Recommendation}: None at this time, this is primarily a technical question with little science impact.


\section{Provide pre-filtered streams}\label{sec:prefilter}

The proposal is to allow brokers to request a pre-filtered alert stream, for example to only include alerts in a certain sky region or which meet other criteria (e.g., brightness limits, number of past detections).

As with the multiple packet formats, the main motivation here is not scientific, but to help reduce brokers' processing costs spent recieving unwanted alerts, or to enable Rubin Observatory to distribute alerts to more brokers.

For example, due to the exponential relationship between the number of variable stars and their variability amplitude, and also that of volume and distance modulus, an apparent magnitude limit that is 1 mag brighter than the nominal 5-sigma DiaSource detection limit could reduce the alert stream data rate by 50\%.

However, all seven full-stream brokers have committed to recieving the full stream, and the term downstream broker refers to brokers which will ingest a filtered stream of alerts from the full-stream brokers.

\textbf{Recommendation}: There is no need for Rubin Observatory to provide pre-filtered streams to brokers.


\section{Distribute delayed alerts as soon as possible}\label{sec:delayed}

The Data Management System (DMS) is required to support the distribution of at least 40,000 alerts per single standard visit, and for visits producing $\leq$40,000 alerts, no more than 1\% of them fail to have at least 98\% of its alerts distributed within 60 seconds of image readout (based on LSR-REQ-0101 in \citeds{LSE-029}; OSS-REQ-0193 in \citeds{LSE-030}; and DMS-REQ-0392 and -0393 in \citeds{LSE-61}).

Furthermore, alert distribution should degrade gracefully beyond that limit, meaning that visits resulting in an excess of 40,000 of alerts should not cause any downtime for the Data Management System (DMS; \citeds{LSE-30}, \citeds{LSE-61}).
It is also a requirement that all alerts be stored in an archival database and be available for retrieval (OSS-REQ-0185 in \citeds{LSE-030}).

This leaves the open question of what, from a science perspective, is the optimal way of dealing with delayed alerts.
(Aspects of the technical implementation of a graceful degradation, such as distributing delayed alerts and alert archive storage access, are out of scope for this document).

There are three main options:

\begin{enumerate}
\item \textbf{Next-opportunity distribution via the alert stream}.
Distribute delayed alerts as soon as possible.
There are plenty of science goals that do not absolutely require alert distribution in 1 minute, and so distributing delayed alerts via the stream would still enable plenty of science.
The brokers might prefer to have delayed alerts clearly flagged to properly process them (e.g., some filtering and processing done by brokers might only be appropriate for alerts delivered within a given latency).
\item \textbf{Next-morning distribution via the alert stream}.
Collect all delayed alerts during the night and then release them (perhaps on a new topic) all at once in the morning, after survey operations have ended for the night.
From a science perspective this is not as useful as next-opportunity distribution, but if it is preferred for technical reasons it would enable more science than the option below.
\item \textbf{Do not distribute delayed alerts; send directly to archive}.
There is no scientific merit in not distributing delayed alerts, and four further drawbacks: the alert archive update timescale is 24 hours (significantly slower than next-morning distribution); the alert database would only be accessible by brokers with authenticated RSP access; alerts might only be able to be queried by alert ID; and bulk download capabilities might be limited.
\end{enumerate}

\textbf{Recommendation}: Flag and distribute delayed alerts as soon as possible to enable time-domain science.

\appendix
% Include all the relevant bib files.
% https://lsst-texmf.lsst.io/lsstdoc.html#bibliographies
\section{References} \label{sec:bib}
\renewcommand{\refname}{} % Suppress default Bibliography section
\bibliography{local,lsst,lsst-dm,refs_ads,refs,books}

% Make sure lsst-texmf/bin/generateAcronyms.py is in your path
\section{Acronyms} \label{sec:acronyms}
\input{acronyms.tex}
% If you want glossary uncomment below -- comment out the two lines above
%\printglossaries





\end{document}
